---
title: "Newnes Mortality Recruitment"
output: html_notebook
---

Model for the effect of mining on plant mortality in Newnes swamps. Code by José R. Ferrer-Paris adapted from David Keith's original.

# Preliminaries

## libraries

```{r}
require(tidyverse)
require(ggeffects)
require(tidyverse)
require(lme4)
```

## Data
object name for newnes mortality in detectable resprouters (i.e. dead remains identifiable) - newnmd

```{r}
newnmd = read.csv("NewnesResprouterMortality_SixDetectableSpp.csv", header=T, row.names=1, stringsAsFactors = FALSE)
```

define variables

```{r}

(newnmd %>% transmute(Val=Valley, Mine=Undermined, Fire=Fire_interval, RecOrgan=Recovery_Organ, Kill, Respr, Mort=Kill/Tot, Sppf=factor(Species)) -> data)

```

# FIRE MORTALITY - DETECTABLE RESPROUTERS

Fit fixed factors first (no random factor)

```{r}
model <- glm(cbind(Kill,Respr)~Mine*Val, data, family=binomial)
summary(model) # see p. 634 Crawley
```
There are many extreme values (Full mortality or full resprouting):

```{r}
hist(data$Mort)
```

Complementary log-log link might be more appropriate in this case, but results are similar for the fixed effect model (not shown):

```{r,eval=FALSE}
model2 <- glm(cbind(Kill,Respr)~Mine*Val, data, family=binomial("cloglog"))
summary(model2) 

```


Check overdispersion - is residual dev >1.5 times greater than residual df? YES it is.

Using a random factor - need a mixed model to deal with random effects of different types of recovery organ. This site has the design & code https://cran.r-project.org/web/packages/ggeffects/vignettes/practical_logisticmixedmodel.html


```{r}
model2 <- glmer(cbind(Kill,Respr) ~ Mine * Val + (1|Sppf), dat=data, family = binomial(link = "logit"))
summary(model2)
plot(model2)
```
Test the complementary-log-log link, some minor differences in residuals, but higher AIC, so we will ignore for now  :

```{r}
model3 <- glmer(cbind(Kill,Respr) ~ Mine * Val + (1|Sppf), dat=data, family = binomial(link = "cloglog"))
AIC(model2,model3)
```

This will print the predicted values and CIs :
```{r}

ggpredict(model2, c("Mine", "Val")) 

```
This will plot the same information:

```{r}
ggpredict(model2, c("Mine", "Val")) %>% plot() 


```
A boxplot of the predictions (based on the logit model):

```{r}
data$prd <- predict(model2,type="response")
boxplot(prd~Val+Mine,data, xlab = "Landform & mining treatment", ylab ="Fire mortality of detectable resprouter plant species",
        cex.axis = 0.85,  names = c("valley floor \n unmined", "valley side \n unmined", "valley floor \n mined", "valley side \n mined"),
        col = c(3,3,7,7))
```


Compare with boxplot of raw values:
```{r}
boxplot(Mort ~ Val + Mine, data, xlab = "Landform & mining treatment", ylab ="Fire mortality (raw values)",
        cex.axis = 0.85,  names = c("valley floor \n unmined", "valley side \n unmined", "valley floor \n mined", "valley side \n mined"),
        col = c(3,3,7,7))

```
