---
title: "Skink counts"
output: html_notebook
---

Model for the effect of mining on skink counts in Newnes swamps. Code by José R. Ferrer-Paris and David Keith

# Preliminaries
## Load required packages

```{r}
require(tidyverse)
require(magrittr)

```

## Read data from CSV file

Data: Newnes swamps soil moisture survey Jan 2013 - Mar 2022 Martin Krogh & Sarsha Gorissen

Read in the data from csv file, which has a header row for col labels and in which character values are not to be converted to factors.

```{r}
skink <- read_csv("NewnSkink.csv")
```

Modify columns in the data frame:

```{r}
skink %<>% mutate(Mine=factor(Mining,levels=c("Unmined reference", "Mined")),Years=Time-2014, Rain=Rainfall_3mth)
```

Visualise data:

```{r}
ggplot(skink, aes(x = SkinkCount, fill = Site)) +
  geom_histogram() + facet_wrap(~ Site)
ggplot(skink, aes(y = SkinkCount, x=Time, colour = Site)) +
  geom_point() + facet_wrap(~ Site)
```

# Linear mixed effects model

Check contrasts for the `Mine` factor:

```{r}
contrasts(newnsoil$Mine)
```

We are including Time since 2014, Mine and Rain in the three months prior to sampling.

We are using a generalized linear model with Poisson error distribution:

```{r}
mdl <- glm(SkinkCount ~ Years * Mine + Rain , skink,family=poisson)
summary(mdl)
```

## Residuals

Check residuals of the model, it looks OK

```{r}
plot(mdl)
```


## Prediction from the model

We calculate model predictions and confidence intervals for the original values of Rain:


```{r}

pred.data <- skink # %>% mutate(Rain=median(Rainfall_3mth))
prd <- predict(mdl,pred.data,type="link",se.fit=T)
ilink <- mdl$family$linkinv
pred.data$prediction <- ilink(prd$fit)
pred.data$pred_upper <- ilink(prd$fit + 1.96 * prd$se.fit)
pred.data$pred_lower <- ilink(prd$fit - 1.96 * prd$se.fit)
```

But we also want to check the prediction confidence interval for the range of rainfall values:

```{r}

pred.data %<>%  mutate(Rain=min(Rainfall_3mth))
prd <- predict(mdl,pred.data,type="link",se.fit=T)
pred.data$pred_min_upper <- ilink(prd$fit + 1.96 * prd$se.fit)
pred.data$pred_min_lower <- ilink(prd$fit - 1.96 * prd$se.fit)

pred.data %<>%  mutate(Rain=max(Rainfall_3mth))
prd <- predict(mdl,pred.data,type="link",se.fit=T)
pred.data$pred_max_lower <- ilink(prd$fit - 1.96 * prd$se.fit)
pred.data$pred_max_upper <- ilink(prd$fit + 1.96 * prd$se.fit)
```


## Figure

First create a basic plot:

```{r}
clrs <- palette()[c(3,7)]
names(clrs) <- c("Unmined reference","Mined")

basic_plot <-
ggplot(skink) + geom_point(aes(y = SkinkCount, x=Time, colour = Mine),size=2) + ggtitle("") +
  xlab("Year") + ylab("Count of Skinks") +
  scale_x_continuous(name="Year", breaks=2015:2022) +
  scale_color_discrete(type = clrs)  + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + guides(colour='none') + facet_wrap(vars(Mine,Site),ncol=3,nrow=2)

```

This is the plot with the prediction and 95% C.I. with the measured values of rain:

```{r}

  basic_plot +
  geom_line(data=pred.data,aes(x=Time,y=prediction)) +  geom_ribbon(data=pred.data,aes(x=Time,y=prediction,ymin=pred_lower,ymax=pred_upper),col=NA,lty=0,alpha=0.3)

```

This is the plot with the prediction and 95% C.I. with the range of values of rain:

```{r}

  basic_plot +
  geom_line(data=pred.data,aes(x=Time,y=prediction)) +  geom_ribbon(data=pred.data,aes(x=Time,y=prediction,ymin=pred_max_lower,ymax=pred_min_upper),col=NA,lty=0,alpha=0.2)

```
